---
title: "01_fundamentals"
author: "Albert Tafur Rangel"
format: html
---

## Design a project

The purpose of this module is create a plot

## Read the data


First download all the datasets to work in this module https://github.com/ae-tafur/data_visualization/tree/main/05_projects/excercises/data

pick up any of the datasets availables and load it. Let the folder in downloads dir. Replace "user" and "your_data" by the name of your user and file.

```{r}

data <- read.csv('C:\User\<username>\Downloads\data\your_data.csv')

```

Now explore the data

```{r}

data

```


Check statistics 

```{r}

summary(data)

```

Select a plot and create one. Here is the code that we have been working on, use it.

```{r}

ggplot(data = DNase) +
  aes(x = conc,
      y = density,
      color = Run) +
  geom_point(alpha = .5,
             # shape = "x",
             stroke = 1,
             size = 4) +
  geom_line() +
  labs(x = "Protein concentration",
       y = "Optical density 400nm",
       title = "ELISA assay for the recombinant protein DNase in rat serum",
       color = "Assay run",
       tag = "a.") +
  theme_minimal()

```

