---
title: "01_fundamentals"
author: "Albert Tafur Rangel"
format: html
---

## Color and visual aesthetics

“Data visualization means drawing graphic displays to show data”

Antony Unwin, 2020.

## Recap

In the last module we end saying: "you can store the plot as a variable"

```{r}

my_plot <- ggplot(data = DNase) +
 aes(x = density,
     y = conc,
     color = Run) +
 geom_point(alpha = .5,
            # shape = "x",
            stroke = 1,
            size = 4) +
 geom_line()

my_plot

```

Now, let's polish the plot. Change the axis name to be more descriptive and add a title.

```{r}

my_plot +
 labs(x = "Optical density 400nm",
      y = "Protein concentration",
      title = "ELISA assay for the recombinant protein DNase in rat serum")

```

We make some improvement. What about change the legend name and add a tag. A tag is very useful in scientific publications to create a panel.

```{r}

my_plot +
 labs(x = "Optical density 400nm",
      y = "Protein concentration",
      title = "ELISA assay for the recombinant protein DNase in rat serum",
      color = "Assay run",
      tag = "a.")

```

Good job! A last adjusment, change the theme. I like too much the minimal theme.

```{r}

my_plot +
 labs(x = "Optical density 400nm",
      y = "Protein concentration",
      title = "ELISA assay for the recombinant protein DNase in rat serum",
      color = "Assay run",
      tag = "a.") +
  theme_minimal()

```

Did it improve? I think yes. But make it sense to have concentration in y-axis? which is the measured variable?

```{r}

ggplot(data = DNase) +
  aes(x = conc,
      y = density,
      color = Run) +
  geom_point(alpha = .5,
             # shape = "x",
             stroke = 1,
             size = 4) +
  geom_line() +
  labs(x = "Protein concentration",
       y = "Optical density 400nm",
       title = "ELISA assay for the recombinant protein DNase in rat serum",
       color = "Assay run",
       tag = "a.") +
  theme_minimal()

```

What a change!

Finally, let's play a little bit with the color. But to use some color palettes a new package need to installed. After, load it.

```{r}

# install.packages("RColorBrewer")
library(RColorBrewer)
  
ggplot(data = DNase) +
  aes(x = conc,
      y = density,
      color = Run) +
  geom_point(alpha = .5,
             # shape = "x",
             stroke = 1,
             size = 4) +
  geom_line() +
  scale_color_brewer(palette = "Spectral") +
  labs(x = "Protein concentration",
       y = "Optical density 400nm",
       title = "ELISA assay for the recombinant protein DNase in rat serum",
       color = "Assay run",
       tag = "a.") +
  theme_minimal()

```

Have you notice that legends are not ordered? Let's adjust it.

```{r}

DNase$Run <- factor(DNase$Run, levels = as.character(1:11))

ggplot(data = DNase) +
  aes(x = conc,
      y = density,
      color = Run)+
  geom_point(alpha = .5,
             # shape = "x",
             stroke = 1,
             size = 4) +
  geom_line() +
  scale_color_brewer(palette = "Spectral") +
  labs(x = "Protein concentration",
       y = "Optical density 400nm",
       title = "ELISA assay for the recombinant protein DNase in rat serum",
       color = "Assay run",
       tag = "a.") +
  theme_minimal()

```

